Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-28 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex

We do not expect any fundamental complications to arise if the poloidal and toroidal magnetic drifts
were retained in the kinetic equation. 
However, we have several reasons for neglecting these magnetic drifts
 in this manuscript.
\begin{enumerate}
\item The ratio of these magnetic drifts to the parallel streaming term we keep is $\sim \rho_* = \rho / L$
 (the gyroradius to scale length), so neglecting the poloidal and toroidal magnetic drifts is
equivalent to considering the $\rho_* \to 0$ limit.
Since our primary interest in this paper is to compare the various options for the electric field
terms and collision operator, it is reasonable to take this $\rho_* \to 0$ limit for simplicity.
\item
Retaining these drifts would make the transport matrix (and figures 1, 2, 4, and 4)
depend on $\rho_*$ in addition to $\nu'$ and $E_*$. 
The dependence of the matrix elements on $\nu'$ and $E_*$ is already  complicated
and interesting to discuss in the $\rho_* \to 0$ limit,
without the extra complexity of finite-$\rho_*$ effects.
%\item Including these magnetic drift terms would also introduce dependence on the radial
%derivatives of the magnetic field
\item The poloidal and toroidal magnetic drifts would add extra complicating
terms to the analysis of moments of the kinetic equation in section III.
%\item If the poloidal and toroidal magnetic drifts are retained without the radial magnetic drift, ????
%\item The poloidal and magnetic drift terms will take additional time to implement in the code, 
%and the 
\end{enumerate}

We have added some text to section II (between (15)-(16)) and a sentence to section V (between (36)-(37)) to clarify these issues.

@@

We have added a new section IV to the paper to discuss momentum balance.
To summarize, we first discuss how parallel momentum has a different status to mass and energy in that mass and
energy are conserved, whereas parallel momentum is not conserved due to the mirror force.
We then compare the $m v_{||}$ moments of the various forms of the drift-kinetic
equation to the full fluid momentum equation, as you suggest.
We show the electric field terms in the kinetic perspective
are associated with a piece of the gyroviscosity in the fluid perspective.  The full trajectory model
gives the correct viscous force whereas the other trajectory models do not.
One further noteworthy property of the parallel velocity moments of the kinetic equations
is that of the trajectory models considered, only the full trajectories preserve intrinsic
ambipolarity in quasisymmetry when $E_r \ne 0$. This result is now discussed at the end of section IV.

\begin{referee}
I do not understand why the source term (25) does not work well with the pitch-angle scattering
operator or P.-A. scattering + momentum conservation operator. It is not the difference in the
collision operator but difference in the trajectory model that is related to the conservation
property of energy, (22). If the original motivation of introducing the term is to maintain the
mass and energy conservation property for finite-Er, then why you need to change the form of
source/sink term if you choose P.-A. scattering operator, which actually conserves the energy,
while you do not need according to the trajectory model? Does it mean diffusion process in
v-coordinate is essential for the source/sink term to work properly?
Please explain the reason of changing the source term according to the collision operator more
detail.
\end{referee}

We have expanded the discussion at the end of section III to clarify this issue.

The system of sources and extra constraint equations serves two independent purposes,
not only to circumvent the conservation problems when $E_r$ is nonzero, but also to eliminate the null
space of the kinetic equation. This latter problem arises even for the DKES trajectories 
and even at $E_r=0$.

When $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints 
can be thought of as a convenient way to keep the linear system square (number of equations = number of unknowns) upon discretization.
Even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate
the null space in the kinetic equation, and the source terms can then be thought of as
a convenient way to keep the linear system square upon discretization.
Any particle and heat source will be sufficient to solve the first problem (conservation), but to solve the second problem (null space),
the number of constraints should match the dimensionality of the null space.  This dimension is set by the collision operator (2 for
Fokker-Planck, $N_x$ for the other two operators.)
This is why the source/constraint scheme is chosen based on the collision operator.

@@

The second part of your question is an interesting observation: when the full trajectory model
is applied to a quasisymmetric field, the source vanishes for any $E_r$, not just at a special ambipolar $E_r$ (as in a non-quasisymmetric field.)
We now state this fact in the new discussion of intrinsic ambipolarity at the end of section IV.

\vspace{0.3in}

In addition to the aforementioned changes, the following modifications have been made to the manuscript:

\begin{itemize}
\item Albert Moll\'{e}n has recently demonstrated precise agreement between SFINCS and an analytical calculation
of Simakov and Helander for the limit of high collisionality [43]. His results are displayed in a new figure 6, with details given in Appendix B,
 and he has been included as a co-author.
\item Output from the DKES code is now plotted in figure 5, to demonstrate the good agreement with SFINCS.
\end{itemize}

### 2014-02-28 doc/sfincsPaper/sfincsPaper.tex
Final version of paper submitted to PoP

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
%\newcommand{\tens}[1]{\overleftrightarrow{#1}}
%\newcommand{\tens}[1]{\overleftrightarrow{\vect{#1}}}
\newcommand{\tens}[1]{\vect{#1}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}
%\newcommand{\changed}[1]{#1}
\newcommand{\changed}[1]{\textcolor{red}{#1}}

@@

In this work, we test the validity of several common simplifying assumptions used in numerical neoclassical calculations
for nonaxisymmetric plasmas.
First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
equation, 
including the commonly used incompressible-$\vect{E}\times\vect{B}$-drift approximation
and two other variants,
corresponding to different effective particle trajectories.
It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible $\vect{E}\times\vect{B}$-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators, including the full linearized Fokker-Planck operator.
At low collisionality, the radial transport driven by radial
gradients is nearly identical for the different operators,
while in other cases it is found to be important that collisions conserve momentum.
%At high collisionality, it seems necessary to use the full Fokker-Planck operator to obtain all transport coefficients correctly.

@@

Another limitation of many past stellarator neoclassical calculations
is that they are often performed with simplified models for collisions.
The linearized Fokker-Planck collision operator -- the most accurate linear operator available -- 
has been implemented in a variety of 
tokamak neoclassical codes \cite{Sauter0, Sauter, WongChan, NEOFP, speedGrids}.
However, due to the numerical challenge of the extra toroidal dimension in stellarators, 
many stellarator neoclassical codes retain only pitch-angle scattering collisions,
so coupling in the energy dimension is eliminated.
The pitch-angle scattering operator
lacks the momentum conservation property of the Fokker-Planck operator,
which is known to be important in many situations \cite{Briesemeister}.
Several techniques have been devised and implemented \cite{Taguchi2, SugamaNishimura, Maassberg, PENTA}
to effectively restore momentum conservation by post-processing
the transport coefficients obtained with a pure pitch-angle scattering operator,
but these methods will not exactly reproduce calculations with the full linearized Fokker-Planck operator.
The NEO-2 code has implemented the full linearized Fokker-Planck operator for
stellarator geometry \cite{Kernbichler1}, but using a field-line-tracing method
which makes it difficult to add the important effect of poloidal $\vect{E}\times\vect{B}$ precession.

Here, we describe a new stellarator neoclassical code
SFINCS (the Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver)
that can be used to explore the aforementioned issues,
comparing various models for effective particle trajectories and collisions.
Although we use the terminology of ``effective trajectories,'' the code
uses continuum rather than Monte Carlo algorithms.
The code solves the 4D drift-kinetic equation for the distribution function,
retaining coupling in 2 spatial independent variables (toroidal and poloidal angle)
and 2 velocity independent variables (speed and pitch angle),
but neglecting radial coupling.
(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, 
while FORTEC-3D \cite{FORTEC_PFR, FORTEC_CPC} is 5D since
radial coupling is retained.)
General nonaxisymmetric nested flux surface geometry is allowed,
one or more species may be included, and several models for collisions are available,
including the full inter-species linearized Fokker-Planck operator.
The incompressible-$\vect{E}\times\vect{B}$-drift trajectories are implemented,
as are several other options for trajectories that include the true $\vect{E}\times\vect{B}$ drift.
As we shall demonstrate, retaining the true form of the $\vect{E}\times\vect{B}$
drift comes at a cost, requiring sources/sinks in the kinetic equation
in order for the solutions to be well behaved.
While all of the various options for the particle trajectories have disadvantages,
SFINCS allows the three options to be compared.
As we will show in several calculations for the LHD and W7-X stellarators,
in many experimentally relevant cases,
the transport matrix elements are nearly identical for the various choices of particle trajectories.
However, differences between the trajectory models emerge when the radial electric field
grows comparable to the ``resonant'' value.

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
\changed{In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
and demonstrate that the electric field terms in the kinetic description
correspond to a component of gyroviscosity in a fluid description.}
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

The aforementioned problems with the partial and full trajectory models may be eliminated 
in the following manner.
The kinetic equation becomes well behaved if we introduce small particle and heat sources
\begin{equation}
S_a(\psi, v) = \Sap(\psi) F_{a}(\psi,v) \left[ x_a^2 - \frac{5}{2}\right] + \Sah(\psi) F_a(\psi, v) \left[x_a^2 - \frac{3}{2}\right]
\label{eq:sources}
\end{equation}
where $\Sap$ and $\Sah$ are considered to be unknowns.
(The factors involving $x_a^2$ in (\ref{eq:sources}) are chosen so $\Sap$ provides a particle source but no heat source, while $\Sah$ provides
a heat source but no particle source.)
As these two new unknowns are now included in the system of equations on each flux surface, we must supply an equal number of additional constraints.
The constraints we supply are $\left< \int d^3v\, f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$,
the sensible requirements that all the flux-surface-averaged density and pressure reside in $F_a$ rather than $f_{a1}$.
This method of modifying the system of equations has the added benefit 
(even in the $d\Phi_0/d\psi=0$ limit)
of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$.
When $S_a$ is included in the kinetic equation, new terms proportional to $\Sap$ and/or $\Sah$
now appear in the mass and energy conservation equations such as (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}).
These conservation equations imply that when $d\Phi_0/d\psi=0$, $\Sap$ and $\Sah$ must vanish.
However, now when $d\Phi_0/d\psi$ is increased from 0 to a small finite number, the sources can turn on
to satisfy  (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}),
eliminating the singular perturbation in $f_{a1}$.  We find that numerical results are then well behaved,
converging appropriately as numerical resolution parameters are increased, and smoothly going to the
$E_r=0$ results as $E_r$ is decreased.

We do not claim that the method proposed here is an ideal solution: the sources (\ref{eq:sources})
are ad-hoc and are not derived rigorously. However, by the techniques proposed here, we can at least
compare the three different trajectory models, and for most experimentally relevant values of $E_r$, we will show
that the three models give nearly identical results. As already mentioned, the need for source terms disappears when the radial electric field equals the value required for ambipolarity, so the source terms do not interfere with the aim of eventually calculating the actual plasma transport.

@@

\changed{
Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = \frac{c}{B} \vect{v}\cdot\vect{b}\times\vect{E} \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different indepedent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, f_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\vect{E} \approx -\nabla \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}
}

@@

Figures (\ref{fig:ErComparison_LHD}) and (\ref{fig:ErComparison_W7X}) show a SFINCS computation of
the ion transport matrix elements for the 3 trajectory models
in two different stellarator geometries.
The calculations in figure (\ref{fig:ErComparison_LHD}) are performed for the $r/a=0.5$ surface of
the LHD stellarator \cite{LHD} in its standard configuration. (Here the flux function $r$ is defined to be proportional to the square root of the toroidal flux enclosed by the flux surface in question.)
The calculations in figure (\ref{fig:ErComparison_W7X}) are performed for the $r/a=0.5$ surface of
the W7-X stellarator \cite{W7X1,W7X2} in its standard configuration.
In the LHD calculation, only the Boozer harmonics of $B(\theta,\zeta)/B_0$ with amplitude larger than $10^{-2}$
are retained, as listed in Table 1 of Ref. \cite{BeidlerBigBenchmarking}.
For both figures, the Fokker-Planck collision operator is used, and the collisionality
is set to $\nu'=0.01$.
As both figures illustrate, the electric field has negligible
effect on the transport matrix elements when $E_* < 0.01$.
For these small values of the electric field, the radial step size for diffusion is 
limited by collisions rather than by $\vect{E}\times\vect{B}$ precession.
As $E_r \to 0$, all the matrix elements converge smoothly to their $E_r=0$ limits.
For $E_*$ in the range $[0.01, 0.3]$,
the $\vect{E}\times\vect{B}$ precession suppresses radial transport,
as can be seen by the reduction in $|L_{11}|$ and $|L_{22}|$. In this regime of $E_*$, the three trajectory models give nearly
identical results for all the transport matrix elements.
However, once $E_*$ exceeds about 0.3, the results from the three trajectory models
begin to separate.

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
%\newcommand{\tens}[1]{\overleftrightarrow{#1}}
%\newcommand{\tens}[1]{\overleftrightarrow{\vect{#1}}}
\newcommand{\tens}[1]{\vect{#1}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}
\newcommand{\changed}[1]{#1}
%\newcommand{\changed}[1]{\textcolor{red}{#1}}

@@

In this work, we examine the validity of several common simplifying assumptions used in numerical neoclassical calculations
for nonaxisymmetric plasmas,
\changed{
both by using a new continuum drift-kinetic code
and by considering analytic properties of the kinetic equation.}
First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
equation, 
including the commonly used incompressible-$\vect{E}\times\vect{B}$-drift approximation
and two other variants,
corresponding to different effective particle trajectories.
It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible $\vect{E}\times\vect{B}$-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators, including the full linearized Fokker-Planck operator.
At low collisionality, the radial transport driven by radial
gradients is nearly identical for the different operators,
while in other cases it is found to be important that collisions conserve momentum.
%At high collisionality, it seems necessary to use the full Fokker-Planck operator to obtain all transport coefficients correctly.

@@

Another limitation of many past stellarator neoclassical calculations
is that they are often performed with simplified models for collisions.
The linearized Fokker-Planck collision operator -- the most accurate linear operator available -- 
has been implemented in a variety of 
tokamak neoclassical codes \cite{Sauter0, Sauter, WongChan, NEOFP, speedGrids}.
However, due to the numerical challenge of the extra dimension in stellarators (i.e., the lack of toroidal symmetry), 
many stellarator neoclassical codes retain only pitch-angle scattering collisions,
so coupling in the energy dimension is eliminated.
The pitch-angle scattering operator
lacks the momentum conservation property of the Fokker-Planck operator,
which is known to be important in many situations \cite{Briesemeister}.
Several techniques have been devised and implemented \cite{Taguchi2, SugamaNishimura, Maassberg, PENTA}
to effectively restore momentum conservation by post-processing
the transport coefficients obtained with a pure pitch-angle scattering operator,
but these methods will not exactly reproduce calculations with the full linearized Fokker-Planck operator.
The NEO-2 code has implemented the full linearized Fokker-Planck operator for
stellarator geometry \cite{Kernbichler1}, but using a field-line-tracing method
which makes it difficult to add the important effect of poloidal $\vect{E}\times\vect{B}$ precession.

Here, we describe a new stellarator neoclassical code
SFINCS (the Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver)
that can be used to explore the aforementioned issues,
comparing various models for effective particle trajectories and collisions.
Although we use the terminology of ``effective trajectories,'' the code
uses continuum rather than Monte Carlo algorithms.
The code solves the 4D drift-kinetic equation for the distribution function,
retaining coupling in 2 spatial independent variables (toroidal and poloidal angle)
and 2 velocity independent variables (speed and pitch angle),
but neglecting radial coupling.
(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, 
while FORTEC-3D \cite{FORTEC_PFR, FORTEC_CPC} is 5D since
radial coupling is retained.)
General nonaxisymmetric nested flux surface geometry is allowed,
one or more species may be included, and several models for collisions are available,
including the full inter-species linearized Fokker-Planck operator.
The incompressible-$\vect{E}\times\vect{B}$-drift trajectories are implemented,
as are several other options for trajectories that include the true $\vect{E}\times\vect{B}$ drift.
As we shall demonstrate, retaining the true form of the $\vect{E}\times\vect{B}$
drift comes at a cost, requiring sources/sinks in the kinetic equation
in order for the solutions to be well behaved.
While all of the various options for the particle trajectories have disadvantages,
SFINCS allows the options to be compared.
As we will show in several calculations for the LHD and W7-X stellarators,
in many experimentally relevant cases,
the transport matrix elements are nearly identical for the various choices of particle trajectories.
However, differences between the trajectory models emerge when the radial electric field
grows comparable to the ``resonant'' value.

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
\changed{In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
and demonstrate that the electric field terms in the kinetic description
correspond to a component of gyroviscosity in a fluid description
only for the most accurate trajectory model.}
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

The aforementioned problems with the partial and full trajectory models may be eliminated 
in the following manner.
The kinetic equation becomes well behaved if we introduce particle and heat sources
\begin{equation}
S_a(\psi, v) = \Sap(\psi) F_{a}(\psi,v) \left[ x_a^2 - \frac{5}{2}\right] + \Sah(\psi) F_a(\psi, v) \left[x_a^2 - \frac{3}{2}\right]
\label{eq:sources}
\end{equation}
where $\Sap$ and $\Sah$ are considered to be unknowns.
(The factors involving $x_a^2$ in (\ref{eq:sources}) are chosen so $\Sap$ provides a particle source but no heat source, while $\Sah$ provides
a heat source but no particle source.)
As these two new unknowns are now included in the system of equations on each flux surface, we must supply an equal number of additional constraints.
The constraints we supply are $\left< \int d^3v\, f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$,
the sensible requirements that all the flux-surface-averaged density and pressure reside in $F_a$ rather than $f_{a1}$.
%This method of modifying the system of equations has the added benefit 
%(even in the $d\Phi_0/d\psi=0$ limit)
%of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$.
When $S_a$ is included in the kinetic equation, new terms proportional to $\Sap$ and/or $\Sah$
now appear in the mass and energy conservation equations such as (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}).
These conservation equations imply that when $d\Phi_0/d\psi=0$, $\Sap$ and $\Sah$ must vanish.
However, now when $d\Phi_0/d\psi$ is increased from 0 to a small finite number, the sources can turn on
to satisfy  (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}),
eliminating the singular perturbation in $f_{a1}$.  We find that numerical results are then well behaved,
converging appropriately as numerical resolution parameters are increased, and smoothly going to the
$E_r=0$ results as $E_r$ is decreased.

We do not claim that the method proposed here is an ideal solution: the sources (\ref{eq:sources})
are ad-hoc and are not derived rigorously. However, by the techniques proposed here, we can at least
compare the three different trajectory models, and for most experimentally relevant values of $E_r$, we will show
that the three models give nearly identical results. 
And as already mentioned, the source terms for the full trajectory model are both zero when the radial electric field equals the value required 
for ambipolarity, so for this model the source terms are really a numerical expedient that
do not affect the transport computations in the end.

@@

\changed{
Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = \frac{c}{B} \vect{v}\cdot\vect{b}\times\vect{E} \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different independent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, f_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\vect{E} \approx -\nabla \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation (\ref{eq:fluidParallelMomentum}) from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}
}

@@

Figures (\ref{fig:ErComparison_LHD}) and (\ref{fig:ErComparison_W7X}) show a SFINCS computation of
the ion transport matrix elements for the 3 trajectory models
in two different stellarator geometries.
The calculations in figure (\ref{fig:ErComparison_LHD}) are performed for the $r/a=0.5$ surface of
the LHD stellarator \cite{LHD} in its standard configuration. (Here the flux function $r$ is defined to be proportional to the square root of the toroidal flux enclosed by the flux surface in question.)
The calculations in figure (\ref{fig:ErComparison_W7X}) are performed for the $r/a=0.5$ surface of
the W7-X stellarator \cite{W7X1,W7X2} in its standard configuration.
In the LHD calculation, only the Boozer harmonics of $B(\theta,\zeta)/B_0$ with amplitude $>10^{-2}$
are retained, as listed in Table 1 of Ref. \cite{BeidlerBigBenchmarking},
whereas all harmonics with relative amplitude $>4\times 10^{-5}$ are retained for the W7-X calculations.
For both figures, the Fokker-Planck collision operator is used, and the collisionality
is set to $\nu'=0.01$.
As both figures illustrate, the electric field has negligible
effect on the transport matrix elements when $E_* < 0.01$.
For these small values of the electric field, the radial step size for diffusion is 
limited by collisions rather than by $\vect{E}\times\vect{B}$ precession.
As $E_r \to 0$, all the matrix elements converge smoothly to their $E_r=0$ limits.
For $E_*$ in the range $[0.01, 0.3]$,
the $\vect{E}\times\vect{B}$ precession suppresses radial transport,
as can be seen by the reduction in $|L_{11}|$ and $|L_{22}|$. In this regime of $E_*$, the three trajectory models give nearly
identical results for all the transport matrix elements.
However, once $E_*$ exceeds about 0.3, the results from the three trajectory models
begin to separate.
